---
title: "bikehamilton-vignette"
output: rmarkdown::html_vignette
abstract: This vignette documents a few different ways that the package and associated dataset can be used, including data analysis. The dataset includes the number of bicycle trips that were recorded in Hamilton through the Transportation Tomorrow Survey.
keywords: bicycling, urban, routing, Canada 
vignette: >
  %\VignetteIndexEntry{bikehamilton-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette illustrates the use of the `bicycle_trips` dataset with some examples of descriptive statistics. The dataset includes the number of bicycle trips that were recorded in Hamilton, as well as the zones of origin and zones of destination. The dataset includes the total number of bicycle trips that occurred between one zone of origin and one zone of destination. These are essentially pairs, namely origin-destination pairs, that generate a certain number of trips (included in the dataset in the `Trips` column).  It documents a few simple ways of manipulating and analyzing the data. The data was collected by the University of Toronto's Data Management Group in the 2016 offering of the Transportation Tomorrow Survey [http://dmg.utoronto.ca/transportation-tomorrow-survey/tts-introduction].

## Load the Packages

Load the packages used in this vignette:
```{r setup}
library(bikehamilton)
library(tidyverse)
```

## Load the Dataset

To load the data, use the function `data()`:
```{r load-data}
data("bicycle_trips")
```

To inspect the dataframe, use the function `summary()`
```{r summary-data}
summary(bicycle_trips)
```

The data frame consists of 3 variables. The variable definitions can be consulted in the help file:
```{r call-help}
?bicycle_trips
```

## What is the average number of bicycle trips per origin-destination pair?

A quick glance at the dataset will reveal that there is a lot of variation in the number of trips that occur between each origin-destination pair. Some origin-destination pairs have over 100 trips, while many have fewer than 20. To calculate the average number of trips, we use the `mean` function:
```{r}
trips = bicycle_trips$Trips    # the number of trips per origin-destination pair 
mean(trips)                    # apply the mean function 
```

## What is the distribution of bicycle trips in Hamilton?

Now that we know the average number of bicycle trips, it is also important to know how the actual number of trips between each origin-destination pair compares to the average number of bicycle trips. We now know that the average number of bicycle trips is approximately 46, but we don't know if most of the origin-destination pairs are below or above the average. To answer this question, we can create a histogram of the `Trips` variable to generate counts:
```{r}
ggplot(data = bicycle_trips) + 
  geom_bar(mapping = aes(x = Trips))
```

We see that most origin-destination pairs produce fewer than 100 trips, with many below the average. This gives us some useful information for further analysis. Why do some origin-destination pairs generate very large number of trips (greater than 200), compared to the average? Why do some origin-destination pairs generate very few trips, less than the average? 
